---
title: "R_S1P_analysis"
author: "Mikhail Dozmorov"
date: "September 6, 2014"
output: html_document
---

```{r setup, echo=FALSE, include=FALSE, cache=FALSE}
# Set up the environment
library(knitr) 
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=T, tidy=T, fig.keep='high', echo=T, dpi=300, out.width=700)
options(replace.assign=TRUE, width=120)
set.seed(1)
```

```{r loadLibraries, echo=FALSE, warning=FALSE, message=FALSE}
suppressMessages(library("GEOquery"))
#library("affy")
suppressMessages(library("limma"))

# Function to get GDS data from GEO, adjust its log2 status and return log2-transformed quantile normalized data
egds <- function(GDS, selected.genes) {
  # Download GDS file, put it in the current directory, and load it. If already downloaded, reuse it
  gds <- getGEO(GDS, destdir="data/", GSEMatrix=T) 
  eset <- GDS2eSet(gds, do.log2=FALSE) # Convert the data to ESET object, do not do any transformation
  # Check if the data is log2-transformed
  ex <- exprs(eset)
  qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
  LogC <- (qx[5] > 100) ||
            (qx[6]-qx[1] > 50 && qx[2] > 0) ||
            (qx[2] > 0 && qx[2] < 1 && qx[4] > 1 && qx[4] < 2)
  if (LogC) { ex[which(ex <= 0)] <- NaN
    exprs(eset) <- log2(ex) } # Log2-transform, if necessary
  exprs(eset) <- normalizeQuantiles(exprs(eset)) # Quantile normalize the data
  # Extract annotation for selected genes
  gpl<-getGEO(Meta(gds)$platform, destdir="data/") # Get the annotation data for this platform from GEO
  annot <- Table(gpl)[Table(gpl)[,"Gene Symbol"] %in% selected.genes, c("ID","Gene Symbol","Gene Title")]
  # Return both objects
  return(list(eset, annot))
}

# Function to get differentially expressed genes based on condition
# Current limitation: hard coded 2-group design in contrast matrix
egds.limma <- function(eset, condition) {
  mod <- model.matrix(formula(paste("~", condition, "+", 0)), data=pData(eset)) # The pData should exist
  colnames(mod) <- group.names
  fit <- lmFit(eset, mod)
  cont.matrix <- makeContrasts(paste(group.names[2], "-", group.names[1]), levels=mod) 
  fit2 <- contrasts.fit(fit, cont.matrix)
  fit2 <- eBayes(fit2, 0.01)
  return(topTable(fit2, number=nrow(fit2$coefficients), adjust.method="BH"))
}

```

We are evaluating ["Molecular Evaluation of Patient-Derived Colorectal Cancer Explants as a Pre-clinical Mouse Model of Colorectal Cancer"](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE35144) study for differential expression of S1P-related genes.

First, we download and pre-process the dataset, and define our genes of interest.

```{r, echo=1:2, message=FALSE, warning=FALSE}
# Select our dataset, and load it into eset Expression set
GDS <- 'GDS4381'
xmap.genes <- read.table("data//genes_mapping.txt", sep="\t", row.names=1) # Genes with cross-mapping
cat(selected.genes <- rownames(xmap.genes))
# Get data from GEO
eset <- egds(GDS, selected.genes)
```

We check phenotype parameters available for this dataset, and specify the one that defines two groups. We check how many samples are in each group.

```{r, echo=1:4}
names(pData(eset[[1]])) # Check which phenoData we have
condition <- "tissue" # Manually specify which one to use, must specify 2 groups
# Check how many samples are in each group
table(pData(eset[[1]])[, condition])
pData(eset[[1]])[, condition] <- make.names(pData(eset[[1]])[, condition]) # Make names R friendly
group.names <- names(table(pData(eset[[1]])[, condition])) # Keep names of the group
```

We get differentially expressed genes, extract fold change and p-values for our genes of interest, and visualize them.

The p-values are adjusted for multiple testing. The fold change (FC) is in regular scale.

Note that most of the genes are assessed by several probes. All probes are visualized, note #####_##_at Affymetrix IDs. Also note that probes having "x" in their name may cross-react, "s" indicate a gene has a single probe.

```{r, echo=FALSE, message=FALSE, fig.height=50, results='hide'}
res <- egds.limma(eset[[1]], condition)
cat("Number of differentially expressed genes:", nrow(res[res$adj.P.Val < 0.1, ]), "out of", nrow(res))
cat("Number of differentially UPregulated genes:", nrow(res[res$adj.P.Val < 0.1 & res$logFC > 0, ]))
cat("Number of differentially DOWNregulated genes:", nrow(res[res$adj.P.Val < 0.1 & res$logFC < 0, ]))
res.selected <- subset(res, Gene.symbol %in% selected.genes)[, c("Gene.symbol", "logFC", "adj.P.Val")]

par(mfrow=c(10,4))
sapply(rownames(res.selected[order(res.selected$Gene.symbol), ]), function(x) {
  boxplot(exprs(eset[[1]])[x, pData(eset[[1]])$tissue == group.names[1]],
          exprs(eset[[1]])[x, pData(eset[[1]])$tissue == group.names[2]],
          main=paste(xmap.genes[as.character(res.selected[x, "Gene.symbol"]), ], x, "\np=", formatC(res.selected[x, "adj.P.Val"], format="e", digits=2), "FC=", formatC(2^res.selected[x, "logFC"], digits=2) ) , xaxt="n", ylab="log expression")
axis(1, at=c(1,2), labels=group.names, las=2)
})
```

```{r, echo=FALSE, eval=FALSE}
# https://stackoverflow.com/questions/14604439/plot-multiple-boxplot-in-one-graph
library(reshape2)
df <- data.frame(t(rbind(pData(eset)$tissue,
            cbind(exprs.selected.q[rownames(res.selected), pData(eset)$tissue == group.names[1]],
                  exprs.selected.q[rownames(res.selected), pData(eset)$tissue == group.names[2]]))))
colnames(df) <- make.names(c("Label", paste(rownames(res.selected), res.selected$Gene)))
row.names(df) <- NULL
df.m <- melt(df, id.var="Label")
require(ggplot2)
ggplot(data = df.m, aes(x=variable, y=value)) + geom_boxplot(aes(fill=Condition))
```


```{r, eval=FALSE, echo=FALSE}
# Use biomaRt for annotation
biocLite("biomaRt") 
library(biomaRt)
listDatasets(useMart("ensembl")) # List available datasets
mart<-useMart("ensembl", dataset="hsapiens_gene_ensembl") # Load BIOMART dataset for homo sapiens
# Information - lists of filters and attributes
head(listFilters(mart), n=50) # Filters, these are our IDs we'll be subsetting the BIOMART annotations on
head(listAttributes(mart), n=50) # Attributes, these are annotations we would like to get for our IDs
Meta(gpl96)$title # Check which microarray we have, to selext the right attributes
attr<-listAttributes(mart) # Get all attributes as a table
attr[grep("affy",attr[,1]),] # Parse them for anything that looks like from affymetrix
# Get annotations from Biomart
genes.up<-getBM(attributes=c('affy_hg_u133a','external_gene_id','description'), filters='affy_hg_u133a', values=up.ids, mart=mart)#, uniqueRows=T)
# Clean workspace
unlink(c("GDS858.soft.gz","genes.up.txt","genes.dn.txt","GPL96.soft"))
```
